A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2021-12-27 analysis/paper/paper.Rmd
Playing around with colours of plots

```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from 19--10 ka (data from Hughes et al. 2016). The study area defined later in the text is marked with a red outline.", out.width = "400px", fig.align="center"}

@@

Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of relative sea-level change at that location, and a DTM edited to remove substantial modern disturbances, the simulations were performed. A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \@ref(fig:example-2)). This year then has a corresponding likely elevation range for the contemporaneous shoreline from which an elevation value was drawn uniformly, using intervals of 5cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing. Polygons were then created from the resulting areas with missing values. The horizontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon. The topographic distance between site and sea was also found by measuring the shortest distance while taking into account the slope of the terrain on the DTM. This was done using the topoDistance package for R [@wang2019]. The topographic distance was measured between site polygon and the horizontally closest point on the shoreline. This means that the distance is not necessarily measured as the closest topographic distance to the shoreline, but rather as the shortest topographic path to the horizontally closest point on the shoreline. This significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact, given the distances considered. The shortest topographic path was found using the Queen's case neighbourhood of eight cells [@conolly2006]. In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygon completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon. While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values were therefore retained, with the exception of for the sites Gunnarsrød 5, Pjonkerød R1 and , where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).

This process was repeated 1000 times for each phase for each site. The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5. Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography. At 1000 simulation runs the analysis of each site ranged from hours to a couple of days, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.

@@

# Create plot showing single simulation of sea-level
seasamp <- ggplot() +
  geom_sf(data = seapoly, fill = "grey", colour = NA) +
  geom_sf(data = sitel, fill = NA, colour = "red") +
  #labs(title =  paste0("Sea-level at ", sealevel, " m above present")) +
  coord_sf(xlim = c(st_bbox(location_bbox)[1], st_bbox(location_bbox)[3]),
           ylim = c(st_bbox(location_bbox)[2], st_bbox(location_bbox)[4]),
           expand = FALSE) +
  ggsn::scalebar(data = sitel, dist = 150, dist_unit = "m",
                 transform = FALSE, st.size = 3, height = 0.3,
                 border.size = 0.1, st.dist = 0.5,
                 anchor = c(x = anc[2] - 500, y = anc[1] + 800)) +
  # Uncomment to annotate with distance measures
  # annotate("text", x = st_bbox(location_bbox)[1] + 20,
  #          y = st_bbox(location_bbox)[4] - 100,
  #          label = dist_label, hjust = 0) +
  theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none")

@@

# Map of simulated sea-levels, date group 1
simsea_plot1 <- shore_plot(output[[1]]$simsea, sitel, 150, date_groups,
                      s_tdist = 0.5, s_xpos = 1300, s_ypos = 800,
                      s_bheight = 0.3, s_tsize = 3)

@@

# Map of simulated sea-levels, date group 2
simsea_plot2 <- shore_plot(output[[2]]$simsea, sitel, 150, date_groups,
                      s_tdist = 0.5, s_xpos = 1300, s_ypos = 800,
                      s_bheight = 0.3, s_tsize = 3)

@@

```{r example-3, echo = FALSE, fig.cap = "The result of 1000 simulation runs for each of the two groups of dates on the site Hegna vest 1. The first column of plots shows the radiocarbon probability density estimate from where dates were drawn during simulation. The second column displays the result of simulating the raised sea-level 1000 times. The darker the colour, the more times the sea-level was simulated to that location. The third column shows violin plots of the different distance measures across all simulations.", out.width = "500px"}
knitr::include_graphics(here("analysis/figures/example_sim_result.png"))
```

@@

The result of running the simulations across all sites are given in Figure \@ref(fig:results). Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, most of the sites have been situated close to the shoreline when they were in use. When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline, as could be expected. To what degree these do in fact represent contamination is discussed further below. However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. It is interesting to note the distance for some of the earliest sites, which appears somewhat high. This, however, can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level. Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, they appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance. This corresponds well with the literature, where the introduction of the Neolithic proper is seen as occurring around this time [@prescott2020; @solheim2021]. With the transition from a predominantly hunter-gatherer lifeway to early agro-pastrolism, settlements in coastal areas are to be increasingly withdrawn from the shoreline.

The negative values around originate from the sites Løvås 1 and 2. These are well-dated sites situated in a relatively undistributed area of the landscape. Thus, while there is a danger of circularity of having archaeological sites inform a reconstruction of the shoreline displacement, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper bound, as they would not have been in use when located beneath sea-level. Furthermore, as the Løvås sites are located at elevations from where the local displacement curve is interpolated between isololation basins at different elevation, it would seem that the sites represent a case where the archaeological material indicates discrepancies in the geological reconstructions of shoreline displacement in the area.

>>>>

```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from c. 17--8 kyr BCE (data from Hughes et al. 2016). The study area defined later in the text is marked with a red outline.", out.width = "400px", fig.align="center"}

@@

Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of relative sea-level change at that location, and a DTM edited to remove substantial modern disturbances, the simulations were performed. A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \@ref(fig:example-2)). This year then has a corresponding likely elevation range for the contemporaneous shoreline from which an elevation value was drawn uniformly, using intervals of 5cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing. Polygons were then created from the resulting areas with missing values. The horizontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon. The topographic distance between site and sea was also found by measuring the shortest distance while taking into account the slope of the terrain on the DTM. This was done using the topoDistance package for R [@wang2019]. The topographic distance was measured between site polygon and the horizontally closest point on the shoreline. This means that the distance is not necessarily measured as the closest topographic distance to the shoreline, but rather as the shortest topographic path to the horizontally closest point on the shoreline. Not finding the topographically closest point significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact on the results, given the distances considered. The shortest topographic path was found using the Queen's pattern neighbourhood of eight cells [@conolly2006]. In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygon completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon. While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values were therefore retained, with the exception of for the sites Gunnarsrød 5, Pjonkerød R1 and , where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).

This process was repeated 1000 times for each phase for each site. The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5. Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography. At 1000 simulation runs the analysis of each site ranged from taking a few hours to a couple of days, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.

@@

# Create plot showing single simulation of sea-level
seasamp <- ggplot() +
  geom_sf(data = seapoly, fill = "#dbe3f3", colour = NA) +
  geom_sf(data = sitel, fill = NA, colour = "red") +
  #labs(title =  paste0("Sea-level at ", sealevel, " m above present")) +
  coord_sf(xlim = c(st_bbox(location_bbox)[1], st_bbox(location_bbox)[3]),
           ylim = c(st_bbox(location_bbox)[2], st_bbox(location_bbox)[4]),
           expand = FALSE) +
  ggsn::scalebar(data = sitel, dist = 150, dist_unit = "m",
                 transform = FALSE, st.size = 3, height = 0.3,
                 border.size = 0.1, st.dist = 0.5,
                 anchor = c(x = anc[2] - 500, y = anc[1] + 800)) +
  # Uncomment to annotate with distance measures
  # annotate("text", x = st_bbox(location_bbox)[1] + 20,
  #          y = st_bbox(location_bbox)[4] - 100,
  #          label = dist_label, hjust = 0) +
  theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none")

@@

# Map of simulated sea-levels, date group 1
simsea_plot1 <- shore_plot(sitearea, output[[1]]$simsea, sitel, 150, date_groups,
                      s_tdist = 0.5, s_xpos = 1300, s_ypos = 800,
                      s_bheight = 0.3, s_tsize = 3)

@@

# Map of simulated sea-levels, date group 2
simsea_plot2 <- shore_plot(sitearea, output[[2]]$simsea, sitel, 150, date_groups,
                      s_tdist = 0.5, s_xpos = 1300, s_ypos = 800,
                      s_bheight = 0.3, s_tsize = 3)

@@

```{r example-3, echo = FALSE, fig.cap = "The result of 1000 simulation runs for each of the two groups of dates on the site Hegna vest 1. The first column of plots shows the radiocarbon probability density estimate from where dates were drawn during simulation. The second column displays the result of simulating the raised sea-level 1000 times. The more opaque the colour, the more times the sea-level was simulated to that location. The third column shows violin plots of the different distance measures across all simulations.", out.width = "500px"}
knitr::include_graphics(here("analysis/figures/example_sim_result.png"))
```

@@

The result of running the simulations across all sites are given in Figure \@ref(fig:results). Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, most of the sites have been situated close to the shoreline when they were in use. When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline, as could be expected. To what degree these do in fact represent contamination is discussed further below. However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. It is interesting to note the distance for some of the earliest sites, which appears somewhat high. This, however, can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level. Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, they appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance. This corresponds well with the literature, where the introduction of the Neolithic proper is seen as occurring around this time [@prescott2020; @solheim2021]. With the transition from a predominantly hunter-gatherer lifeway to early agro-pastoralism, settlements in coastal areas are to be increasingly withdrawn from the shoreline.

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. These are recently excavated, well-dated sites situated in a relatively undistributed area of the landscape. Thus, while there is a danger of circularity of having archaeological sites inform a reconstruction of the shoreline displacement, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit as they would not have been in use when located beneath sea-level. Furthermore, as the Løvås sites are located at elevations from where the local displacement curve is interpolated between isololation basins at different elevation, it would seem that the sites represent a case where the archaeological material indicates discrepancies in the geological reconstructions of shoreline displacement in the area.
